Simple Models of Metal-Line Absorption and Emission from Cool Gas Outflows

### 2010-11-17 Winds/papers/Theory/First/Submit/ms.tex
profx
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5413 ac11df5e-a788-c23b-a8e1-c758c323924d

\title{Simple Wind Models of Cool Gas Outflows as Viewed by
  Metal-Line Absorption and Emission}

@@

\begin{abstract}
We analyze the absorption and emission-line profiles produced by a set
of simple, cool gas wind models motivated by galactic-scale outflow
observations.  
We implement monte carlo radiative
transfer techniques that track the propagation of scattered and
fluorescent photons to generate 1D spectra and 2D spectral images.
{\bf [Should we make the codes publicly available?  Upon request?]}
%We examine outflows with a range of power-law density and velocity
%profiles and a radiative-driven wind model.  
We focus on the \mgiid\ doublet and \ion{Fe}{2} UV1
multiplet at $\lambda \approx 2600$\AA, but the results are applicable
to other transitions that trace outflows (e.g.\ \ion{Na}{1},
\ion{H}{1} \lya, \ion{Si}{2}).   
By design, the resonance transitions show blue-shifted
absorption but one also predicts strong resonance and fine-structure line-emission at roughly
the systemic velocity.  This line-emission `fills-in' the absorption
reducing the equivalent width by up to $50\%$, shifting the centroid
by tens of \kms, and severely
reducing the effective
opacity near systemic.  Analysis of cool gas outflows that ignores this
line-emission may incorrectly infer that the gas is partially
covered, measure a significantly lower peak optical depth, and/or
conclude that gas at systemic velocity is absent   
(e.g.\ an interstellar or slowly infalling component). 
Because the \ion{Fe}{2} lines
are connected by optically-thin transitions to
fine-structure levels, their profiles
more closely reproduce the intrinsic opacity
of the wind.   Together these results naturally explain the
absorption and emission-line characteristics observed for star-forming
galaxies at $z<1$.
We also study a scenario purported to describe the outflows
of $z \sim 3$ Lyman break galaxies and argue this model fails to
reproduce the observations when radiative transfer of scattered photons
is properly considered.   
%Strong line-emission from metal-line
%transitions is a generic prediction of wind models, even in the
%presence of dust and anisotropic configurations.  
Although line-emission complicates the analysis of absorption-line
profiles, the 
surface brightness profiles 
offer a unique means of assessing the morphology and size
of galactic-scale winds.  Furthermore, the kinematics and line-ratios
offer powerful diagnostics of outflows,  motivating deep,
spatially-extended spectroscopic observations.

@@

\keywords{absorption lines -- intergalactic medium -- Lyman limit systems -- SDSS}

@@

Nearly all gaseous objects that shine are also
observed to generate gaseous flows.  This includes the jets of protostars, the
stellar winds of massive O and B stars, the gentle Solar wind
of our Sun, the associated absorption of bright quasars, and the
spectacular jets of radio-loud AGN.   These gaseous outflows 
%%KHRR
regulate the metal and dust content and distribution within the 
objects and their surroundings, 
%%
moderate the accretion of new material, and %onto the object, 
inject energy and momentum into gas on large scales. %, and ...
Developing a comprehensive model for these flows is critical to
understanding the evolution of the source and its impact on the
surrounding environment.

Starburst galaxies, whose luminosity is dominated by \ion{H}{2} regions
and massive stars, are also observed to drive gaseous outflows.  These
flows are generally expected (and sometimes observed) to have multiple
phases, for example a hot and diffuse phase traced by X-ray emission
together with a cool, denser phase traced by H$\alpha$ emission 
%% KHRR -- added refs
\citep[e.g.][]{ham90,martin99,shc+04,km10}. 
Several spectacular
examples in the local universe demonstrate that flows can extend to
up to $\sim 10$ kpc
%tens of kpc 
from the galaxy \citep{lhw99,vsr+03,wsg08} carrying significant speed
to escape from the gravitational potential well of the galaxy's dark
matter halo \citep[e.g.][]{sh09}.

%Beyond $z \sim 0$, 
Galactic outflows are also
revealed by UV and optical absorption lines, e.g.\ \ion{Na}{1},
\ion{Mg}{2}, \ion{Si}{2} and \ion{C}{4} transitions.  With the galaxy
as a backlight, one observes gas that is predominantly
blue-shifted which indicates a flow toward
Earth and away from the galaxy.  These transitions are sensitive to
the cool (\ion{Mg}{2}; $T \sim 10^4$K) and warm (\ion{C}{4}; $T \sim
10^5$K) phases of the flow.  
The incidence of cool gas outflows is
nearly universal in 
%% KHRR added:
vigorously
%%
star-forming galaxies;  this includes systems at $z \sim 0$
which exhibit \ion{Na}{1} and \ion{Mg}{2} absorption
\citep{rvs05a,martin05,smn+09,mb09,cth+10}, 
$z \sim 1$ star-forming galaxies traced by
\ion{Fe}{2} and \ion{Mg}{2} transitions \citep{wcp+09,rwk+10}, and
$z>2$ Lyman break galaxies (LBGs) that show blue-shifted \ion{Si}{2},
\ion{C}{2}, and \ion{O}{1} transitions \citep{sgp+96,lkg+97,shapley03}.

@@

Bright \ion{Mg}{2} line emission has also been
reported for individual galaxies at $z \sim 0.7$ by \cite[][see also
Rubin et al\ 2011, in prep.]{rubin+10c}.
In their analysis of a single galaxy spectrum, \cite{rubin+10c}
further demonstrated that the \ion{Mg}{2} line emission is spatially
extended, and used the size of the emission to infer that the wind 
extends to at least 7\,kpc from the galaxy.
These authors additionally detected line
emission from non-resonant \ion{Fe}{2}$^*$ transitions, and attributed
the emission to fluorescence powered by \ion{Fe}{2} resonant
absorption.  In other words, these photons are re-emitted by the wind
into our sightline, and are analogous to the emitted photons in a P-Cygni profile.
%Their interpretation is that the emission is the result of photons
%scattered off the wind into our sightline, analogous to P-Cygni.
%\cite{rubin+10c} also observed spatially extended \ion{Mg}{2} emission
%which they use to estimate the size of the outflow.  They infer that
%the wind extends to at least 7\,kpc from the galaxy.  
Line-emission is
also observed for $z \sim 3$ LBGs in the resonant \lya\ transition
and non-resonant \ion{Si}{2}$^*$ transitions\footnote{See also
  \cite{france10} for a new example at $z \sim 0$.} \citep{prs+02,shapley03}.
We emphasize that this emission arises from a completely different
physical process than those generating X-ray and H$\alpha$ emission
(e.g., shocks), and presumably probes both the cool gas at the base of
the wind and the outskirts of the flow (i.e., wherever a given
transition is optically thick).
A comprehensive analysis of the %emission lines
scattered and fluorescent emission
related to galactic-scale outflows
(e.g.\ via deep integral-field-unit [IFU] observations,
\citealt{ssb+05,wsg08}) 
may offer unique
diagnostics on the spatial extent, morphology, and density of the outflow 
from distant galaxies,
eventually
setting tighter constraints on the energetics of the flow.

@@

In this paper, we focus on two sets of radiative transitions
arising from Fe$^+$ and Mg$^+$ ions
(Table~\ref{tab:atomic}, Figure~\ref{fig:energy}).
This is a necessarily limited
set, but the two ions and their transitions do have characteristics
shared by the majority of low-ion transitions
observed in cool-gas outflows. Therefore, many
of the results that follow may be generalized to observational studies that
consider other atoms and ions of cool gas.

@@

Our calculations also ignore collisional
processes\footnote{Recombination is also ignored.}, i.e.\ collisional
excitation and de-excitation of the various levels.  For the
fine-structure levels of the \aconfig\ configuration, the excitation
energies are modest ($T_{\rm ex} \sim 1000$\,K) but the critical
density $n_c$ is large.  For the \aconfig$_{9/2} \to
$\aconfig$_{7/2}$ transition, the critical density $n_e^C \approx 4
\sci{5} \cm{-3}$.  At these densities, one would predict  
detectable quantities of \ion{Fe}{1} which has not yet been observed
in galactic-scale outflows. Furthermore, observations rarely show
{\it absorption} from the
fine-structure levels of the \aconfig\ configuration and that material
is not significantly blue-shifted (Rubin et al., in prep). 
In the following, we assume that electrons only occupy the
ground-state, i.e.\ the gas has zero opacity to the non-resonant
lines.  If collisional excitation is insignificant 
then one may also neglect collisional de-excitation.  
Regarding the \ion{Mg}{2} doublet, its
excitation energy is significantly higher implying 
negligible collisional processes at essentially any density.

@@

Nearly all of the absorption studies of galactic-scale outflows 
have focused on intensely, star-forming galaxies.  The
intrinsic emission of these galaxies is a complex combination of
stars and \ion{H}{2} regions that is then modulated by dust and gas
within the interstellar medium (ISM).  For the spectral regions studied
here, most stars show a featureless continuum, but a few spectral
types do show significant \ion{Mg}{2} and \ion{Fe}{2} absorption.
In addition, asymmetric and/or blue-shifted absorption is exhibited in these
transitions in A and F stars driving stellar winds \citep{sll+94}.
\ion{Mg}{2} P Cygni profiles are observed in a handful of F stars by \cite{sll+94}, who
attribute the emission to chromospheric activity rather than mass-loss effects.
\ion{H}{2} regions, meanwhile, are observed to emit weakly
at the \mgiid\ doublet, primarily due to recombinations in the outer
layers \citep{kbc+93}.
It is beyond the scope of this paper to
properly model the 
stellar absorption and \ion{H}{2} region emission, but the reader
should be aware that they can complicate the observed spectrum,
independently of any outflow, especially at velocities $v \approx 0 \mkms$.
In the following, we assume a simple flat continuum 
normalized to unit value.  The size of the emitting
region $r_{\rm source}$ is a free parameter, but we restrict its value
to be smaller than the minimum radial extent of any gaseous
component.   Lastly, the source does not absorb any scattered or
emitted photons.

@@

To validate the monte carlo code, we calculated a series of test problems consisting of a point source embedded in a spherical, homogenous medium of varying resonant line optical depth.  The resulting resonant line profiles  \citep[see][]{Kasen_lyman} were
found to be in good agreement with the standard analytic solutions
\citep{Harrington_1973,Neufeld_1990,Dijkstra_2006}.  As a further
test, we calculated line profiles for the case of an extended
spherical source embedded in a homologously expanding wind with large
velocity gradient.  These model spectra
(Figure~\ref{fig:oneline_test}) are in good agreement with ones
determined by direct numerical integration of the radiation transport
equation under the Sobolev approximation \citep[e.g.,][]{Jeffery_Branch}.

@@

A photon propagating through a differentially expanding medium
interacts with a line only when
its comoving frame wavelength is Doppler-shifted into near resonance with the line center rest wavelength. 
For a wind with a steep velocity law  (i.e.\ a large gradient $dv/dr$)
and/or a narrow intrinsic profile (i.e.\ a small Doppler parameter), the 
spatial extent of the region of resonance
may be much smaller than the length scale of the
wind itself.  The interaction can then 
be considered to occur at  a point.
In this case, \citet{sobolev60} introduced a formalism that gives the line optical depth $\tau(r)$ at  
a given point in terms of the density and
velocity gradient of the flow at that radius.  For a wind in homologous expansion,
this optical depth is independent of the direction of propagation and
is given by
\begin{equation}
\tau_S(r) = K_\ell   \lambda^0_\ell | dv/dr |^{-1} \cmma
\label{eqn:Sobolev}
\end{equation}
where 
\begin{equation}
K_\ell = \frac{\pi e^2}{m_e c} f_\ell n_\ell(r)
\end{equation}
is the integrated line opacity, $\lambda^0_\ell$ the rest center wavelength,
$f_\ell$ the oscillator strength and
$n_\ell$ the density in the lower level of the transition.   We have
neglected corrections for stimulated emission.   This optical
depth applies to a photon with wavelength $\lambda_S
\equiv \lambda_\ell (1- v(r)~\hat{r}  \cdot \hat{d} /c)$ where $\hat{d}$ is the 
direction of propagation.  The probability that such a photon
is scattered/absorbed at the point of resonance is simply $1 - \exp[-\tau_S(r)]$.
We find that the Sobolev approximation applies for nearly all of the
models presented in this paper, and therefore provides a convenient
approach to estimating the optical depth.

@@

Using the methodology described in $\S$~\ref{sec:method}, we
propagated photons from the source and through the outflow to an
`observer' at $r \gg r_{\rm outer}$ who views the entire wind+source
complex.  Figure~\ref{fig:fiducial_1d} presents the 1D spectrum
that this observer would record, with the unattenuated flux
normalized to unit value.   The \ion{Mg}{2} doublet
shows the canonical `P-Cygni profile' that characterizes a continuum
source embedded within an outflow.  Strong absorption is evident at
$\delta v  \equiv (\lambda/\lambda_0 - 1) < -50 \mkms$ in both
transitions (with equivalent widths $W^{2796} =
3.0$\AA\ and $W^{2803} = 1.3$\AA) and each shows emission at
positive velocities.  For an isotropic and dust-free model, the
total equivalent width of the doublet must be zero,
i.e.\ every photon
absorbed eventually escapes the system, typically at lower
energy.  The wind simply shuffles the photons in frequency space.
A simple summation of the absorption and emission equivalent widths
(Table~\ref{tab:fiducial_EW}) confirms this expectation.

Focusing further on the \ion{Mg}{2} absorption, one notes that the profiles lie
well above zero intensity and have similar depth even though their $f\lambda$
values differ by a factor of two.  In standard absorption-line
analysis, this is 
the tell-tale signature of a `cloud' that has a high optical depth (i.e.\
saturated) which only partially covers the emitting source
\citep[e.g.][]{rvs05a,hkp+10}.  Our fiducial wind model, however, 
{\it entirely covers the source}; the apparent partial covering must
be related to a different effect.
Figure~\ref{fig:noemiss} further emphasizes this point by comparing the 
absorption profiles from Figure~\ref{fig:fiducial_1d} against an
artificial model where no absorbed photons are 
re-emitted.   As expected from the
$\tau_{2796}$ profile (Figure~\ref{fig:fiducial_nvt}), this
`intrinsic' model
produces a strong \mgiid\ doublet that absorbs all photons at
$\delta v \approx -100 \mkms$, i.e.\ $I_{2796}^{\rm min} = \exp(-\tau^{\rm
  max}_{2796}) \approx 0$.
The true model, in contrast, has been `filled in' at $\delta v \approx -100
\mkms$ by photons scattered in the wind.  An
absorption-line analysis that ignores these effects
would (i) systematically underestimate the true optical
depth and/or (ii) falsely conclude that the wind partially covers the
source.  We will find that these are
generic results, even for wind models that include
dust and are not fully isotropic.

@@

Now consider the \ion{Fe}{2} transitions:
the bottom left panel of Figure~\ref{fig:fiducial_1d} covers the
majority of the \ion{Fe}{2} UV1 transitions and several are
shown in the velocity plot.  The line
profile for \feiib\ is very similar to the \mgiid\ doublet;
one observes strong absorption to negative velocities and strong
emission at $\mdv > 0 \mkms$ producing a characteristic P-Cygni profile. 
Splitting
the profile at $\mdv = -50 \mkms$, we measure an equivalent width
$W^{2600}_{\rm a} = 1.16$\AA\ in absorption and $W^{2600}_{\rm e} =
-0.94$\AA\ in emission (Table~\ref{tab:fiducial_EW}) for a total
equivalent width of $W^{2600}_{\rm TOT} \approx 0.22$\AA.  
In contrast, the \feiia\ resonance line shows much weaker emission and
a much higher total equivalent width ($W^{2586}_{\rm TOT} = 0.49$\AA),
even though the line has a $2 \times$ lower $f\lambda$ value.
These differences between the \ion{Fe}{2} resonance lines (and with the
\ion{Mg}{2} doublet) occur because of the complex of non-resonant
\feiis\ transitions that are coupled to the resonance lines
(Figure~\ref{fig:energy}).  Specifically, 
a resonance photon absorbed at \feiid\ has a finite probability of
being re-emitted as a non-resonant photon which then escapes the system
without further interaction.  The principal effects are to reduce the
line emission of \feiid\ and to produce non-resonant line-emission (e.g.\
\feiic).

@@

The preceding discussion emphasizes the filling-in of resonance
absorption at $\mdv
\lesssim -50 \mkms$ and the generation of emission lines at $\mdv \approx
0 \mkms$ by photons scattered in the wind.  To study the spatial
extent of this emission, we mapped the emission of the fiducial
model (see $\S$~\ref{sec:monte} for a description of the algorithm).
The output is a set of surface-brightness maps in a series of 
frequency channels yielding a
dataset analogous to integral-field-unit (IFU) observations.  In
Figure~\ref{fig:fiducial_ifu_mgii}, we present the output 
at several velocities relative to the \mgiia\
transition. At $\mdv = -250 \mkms$, where the wind has an optical
depth $\tau_{2796} < 1$ (Figure~\ref{fig:fiducial_nvt}),
the source contributes roughly half of the observed flux.  
At $\mdv=-100 \mkms$, however, the
wind absorbs all photons from the source and the observed emission is
entirely from photons scattered by the wind.  This scattered emission
actually exceeds the source+wind emission at 
$\mdv = -250 \mkms$ such that the absorption profile is
negatively offset from the velocity where $\tau_{2796}$ is maximal
(Figure~\ref{fig:fiducial_nvt}).
The net result is
weaker \ion{Mg}{2} absorption that peaks blueward of the actual peak in the
optical depth profile.  
Table~\ref{tab:line_diag} reports several kinematic measurements of
the absorption and emission features.
Clearly, these effects complicate estimates for the
speed, covering fraction, and total column density of the wind.  At
$\mdv = 0 \mkms$, the wind and source have comparable total flux with the
latter dominating at higher velocities.

Similar results are observed for the \ion{Fe}{2} resonance
transitions (Figure~\ref{fig:fiducial_ifu_feii}).
For transitions to fine-structure levels of the \aconfig\
configuration, the source
is unattenuated but there is a significant contribution from photons
generated in the wind. 
The results presented in Figures~\ref{fig:fiducial_ifu_mgii} and
\ref{fig:fiducial_ifu_feii} are sensitive to the radial extent,
morphology, density and velocity profiles of this galactic-scale
wind.  Consequently, IFU observations of line emission from low-ion
transitions may offer the most direct constraints on galactic-scale
wind properties.

>>>>

\title{Simple Models of Metal-Line Absorption and Emission from
Cool Gas Outflows}

@@

\begin{abstract}
We analyze the absorption and emission-line profiles produced by a set
of simple, cool gas wind models motivated by galactic-scale outflow
observations.  
We implement monte carlo radiative
transfer techniques that track the propagation of scattered and
fluorescent photons to generate 1D spectra and 2D spectral images.
{\bf [Should we make the codes publicly available?  Upon request?]}
%We examine outflows with a range of power-law density and velocity
%profiles and a radiative-driven wind model.  
We focus on the \mgiid\ doublet and \ion{Fe}{2} UV1
multiplet at $\lambda \approx 2600$\AA, but the results are applicable
to other transitions that trace outflows (e.g.\ \ion{Na}{1},
\ion{H}{1} \lya, \ion{Si}{2}).   
By design, the resonance transitions show blue-shifted
absorption but one also predicts strong resonance and fine-structure line-emission at roughly
the systemic velocity.  This line-emission `fills-in' the absorption
reducing the equivalent width by up to $50\%$, shifting the centroid
by tens of \kms, and severely
reducing the effective
opacity near systemic.  Analysis of cool gas outflows that ignores this
line-emission may incorrectly infer that the gas is partially
covered, measure a significantly lower peak optical depth, and/or
conclude that gas at systemic velocity is absent   
(e.g.\ an interstellar or slowly infalling component). 
Because the \ion{Fe}{2} lines
are connected by optically-thin transitions to
fine-structure levels, their profiles
more closely reproduce the intrinsic opacity
of the wind.   Together these results naturally explain the
absorption and emission-line characteristics observed for star-forming
galaxies at $z<1$.
We also study a scenario promoted to describe the outflows
of $z \sim 3$ Lyman break galaxies and argue this model fails to
reproduce the observations when radiative transfer of scattered photons
is properly considered.   
%Strong line-emission from metal-line
%transitions is a generic prediction of wind models, even in the
%presence of dust and anisotropic configurations.  
Although line-emission complicates the analysis of absorption-line
profiles, the 
surface brightness profiles 
offer a unique means of assessing the morphology and size
of galactic-scale winds.  Furthermore, the kinematics and line-ratios
offer powerful diagnostics of outflows,  motivating deep,
spatially-extended spectroscopic observations.

@@

\keywords{galaxies: formation -- galaxies: starbust}

@@

Nearly all gaseous objects that shine are also
observed to generate gaseous flows.  This includes the jets of protostars, the
stellar winds of massive O and B stars, the gentle Solar wind
of our Sun, the associated absorption of bright quasars, and the
spectacular jets of radio-loud AGN.   These gaseous outflows 
regulate the metal and dust content and distribution within the 
objects and their surroundings, 
moderate the accretion of new material, and %onto the object, 
inject energy and momentum into gas on large scales. %, and ...
Developing a comprehensive model for these flows is critical to
understanding the evolution of the source and its impact on the
surrounding environment.

Starburst galaxies, whose luminosity is dominated by \ion{H}{2} regions
and massive stars, are also observed to drive gaseous outflows.  These
flows are generally expected (and sometimes observed) to have multiple
phases, for example a hot and diffuse phase traced by X-ray emission
together with a cool, denser phase traced by H$\alpha$ emission 
\citep[e.g.][]{ham90,martin99,shc+04,km10}. 
Several spectacular
examples in the local universe demonstrate that flows can extend to
up to $\sim 10$ kpc
from the galaxy \citep{lhw99,vsr+03,wsg08} carrying significant speed
to escape from the gravitational potential well of the galaxy's dark
matter halo \citep[e.g.][]{sh09}.

Galactic outflows are also
revealed by UV and optical absorption lines, e.g.\ \ion{Na}{1},
\ion{Mg}{2}, \ion{Si}{2} and \ion{C}{4} transitions.  With the galaxy
as a backlight, one observes gas that is predominantly
blue-shifted which indicates a flow toward
Earth and away from the galaxy.  These transitions are sensitive to
the cool (\ion{Mg}{2}; $T \sim 10^4$K) and warm (\ion{C}{4}; $T \sim
10^5$K) phases of the flow.  
The incidence of cool gas outflows is
nearly universal in vigorously
star-forming galaxies;  this includes systems at low $z$
which exhibit \ion{Na}{1} and \ion{Mg}{2} absorption
\citep{rvs05a,martin05,smn+09,mb09,cth+10}, 
$z \sim 1$ star-forming galaxies with winds traced by
\ion{Fe}{2} and \ion{Mg}{2} transitions \citep{wcp+09,rwk+10}, and
$z>2$ Lyman break galaxies (LBGs) that show blue-shifted \ion{Si}{2},
\ion{C}{2}, and \ion{O}{1} transitions \citep{sgp+96,lkg+97,shapley03}.

@@

Bright \ion{Mg}{2} line emission has also been
reported for individual galaxies at $z \sim 0.7$ by \cite[][see also
Rubin et al\ 2011, in prep.]{rubin+10c}.
In their analysis of a single galaxy spectrum, \cite{rubin+10c}
further demonstrated that the \ion{Mg}{2} line emission is spatially
extended, and used the size of the emission to infer that the wind 
extends to at least 7\,kpc from the galaxy.
These authors additionally detected line
emission from non-resonant \ion{Fe}{2}$^*$ transitions, and attributed
the emission to fluorescence powered by \ion{Fe}{2} resonant
absorption.  In other words, these photons are re-emitted by the wind
into our sightline, and are analogous to the emitted photons in a P-Cygni profile.
%Their interpretation is that the emission is the result of photons
%scattered off the wind into our sightline, analogous to P-Cygni.
%\cite{rubin+10c} also observed spatially extended \ion{Mg}{2} emission
%which they use to estimate the size of the outflow.  They infer that
%the wind extends to at least 7\,kpc from the galaxy.  
Line-emission that may be related to outflows is
also observed for $z \sim 3$ LBGs in the resonant \lya\ transition
and non-resonant \ion{Si}{2}$^*$ transitions\footnote{See also
  \cite{france10} for a new example at $z \sim 0$.} \citep{prs+02,shapley03}.
This emission likely arises from a completely different
physical process than those generating X-ray and H$\alpha$ emission
(e.g., shocks), and presumably probes both the cool gas at the base of
the wind and the outskirts of the flow (i.e., wherever a given
transition is optically thick).
A comprehensive analysis of the %emission lines
scattered and fluorescent emission
related to galactic-scale outflows
(e.g.\ via deep integral-field-unit [IFU] observations,
\citealt{ssb+05,wsg08}) 
may offer unique
diagnostics on the spatial extent, morphology, and density of the outflow 
from distant galaxies,
eventually
setting tighter constraints on the energetics of the flow.

@@

In this paper, we focus on two sets of radiative transitions
arising from Fe$^+$ and Mg$^+$ ions
(Table~\ref{tab:atomic}, Figure~\ref{fig:energy}).
This is a necessarily limited
set, but the two ions and their transitions do have characteristics
shared by the majority of low-ion transitions
observed in cool-gas outflows. Therefore, many
of the results that follow may be generalized to observational studies that
consider other atoms and ions tracing cool gas.

@@

Our calculations also ignore collisional
processes\footnote{Recombination is also ignored.}, i.e.\ collisional
excitation and de-excitation of the various levels.  For the
fine-structure levels of the \aconfig\ configuration, the excitation
energies are modest ($T_{\rm ex} \sim 1000$\,K) but the critical
density $n_c$ is large.  For the \aconfig$_{9/2} \to
$\aconfig$_{7/2}$ transition, the critical density $n_e^C \approx 4
\sci{5} \cm{-3}$.  At these densities, one would predict  
detectable quantities of \ion{Fe}{1} which has not yet been observed
in galactic-scale outflows. 
If collisional excitation is insignificant 
then one may also neglect collisional de-excitation.  
Furthermore, observations rarely show
{\it absorption} from the
fine-structure levels of the \aconfig\ configuration and that material
is not significantly blue-shifted (Rubin et al., in prep). 
In the following, we assume that electrons only occupy the
ground-state, i.e.\ the gas has zero opacity to the non-resonant
lines.  
Regarding the \ion{Mg}{2} doublet, its
excitation energy is significantly higher implying 
negligible collisional processes at essentially any density.

@@

Nearly all of the absorption studies of galactic-scale outflows 
have focused on intensely, star-forming galaxies.  The
intrinsic emission of these galaxies is a complex combination of
light from stars and \ion{H}{2} regions that is then modulated by dust and gas
within the ISM.  For the spectral regions studied
here, the hottest stars show a featureless continuum, but later spectral
types do show significant \ion{Mg}{2} and \ion{Fe}{2} absorption.
In addition, asymmetric and/or blue-shifted absorption is exhibited in these
transitions in A and F stars driving stellar winds \citep{sll+94}.
\ion{Mg}{2} P Cygni profiles are observed in a handful of F stars by \cite{sll+94}, who
attribute the emission to chromospheric activity rather than mass-loss effects.
\ion{H}{2} regions, meanwhile, are observed to emit weakly
at the \mgiid\ doublet, primarily due to recombinations in the outer
layers \citep{kbc+93}.
It is beyond the scope of this paper to
properly model the 
stellar absorption and \ion{H}{2} region emission, but the reader
should be aware that they can complicate the observed spectrum,
independently of any outflow, especially at velocities $v \approx 0 \mkms$.
In the following, we assume a simple flat continuum 
normalized to unit value.  The size of the emitting
region $r_{\rm source}$ is a free parameter, but we restrict its value
to be smaller than the minimum radial extent of any gaseous
component.   Lastly, the source does not absorb any scattered or
emitted photons.

@@

To validate the monte carlo code, we calculated a series of test problems consisting of a point source embedded in a spherical, homogenous medium of varying resonant line optical depth.  The resulting resonant line profiles  \citep[see][]{Kasen_lyman} were
found to be in good agreement with the standard analytic solutions
\citep{Harrington_1973,Neufeld_1990,Dijkstra_2006}.  As a further
test, we calculated line profiles for the case of an extended
spherical source embedded in a homologously expanding wind with large
velocity gradient.  These model spectra
(Figure~\ref{fig:oneline_test}) are in good agreement with ones
determined by direct numerical integration of the radiation transport
equation under the Sobolev approximation
\citep[$\S$~\ref{sec:Sobolev}; e.g.,][]{Jeffery_Branch}.

@@

A photon propagating through a differentially expanding medium
interacts with a line only when
its comoving frame wavelength is Doppler-shifted into near resonance with the line center rest wavelength. 
For a wind with a steep velocity law  (i.e.\ a large gradient $dv/dr$)
and/or a narrow intrinsic profile (i.e.\ a small Doppler parameter), the 
spatial extent of the region of resonance
may be much smaller than the length scale of the
wind itself.  The interaction can then 
be considered to occur at  a point.
In this case, \citet{sobolev60} introduced a formalism that gives the line optical depth $\tau(r)$ at  
a given point in terms of the density and
velocity gradient of the flow at that radius.  For a wind in homologous expansion,
this optical depth is independent of the direction of propagation and
is given by
\begin{equation}
\tau_S(r) = K_\ell   \lambda^0_\ell | dv/dr |^{-1} \cmma
\label{eqn:Sobolev}
\end{equation}
where 
\begin{equation}
K_\ell = \frac{\pi e^2}{m_e c} f_\ell n_\ell(r)
\end{equation}
is the integrated line opacity, $\lambda^0_\ell$ the line-center rest wavelength,
$f_\ell$ the oscillator strength and
$n_\ell$ the density in the lower level of the transition.   We have
neglected corrections for stimulated emission.   This optical
depth applies to a photon with wavelength $\lambda_S
\equiv \lambda_\ell (1- v(r)~\hat{r}  \cdot \hat{d} /c)$ where $\hat{d}$ is the 
direction of propagation.  The probability that such a photon
is scattered/absorbed at the point of resonance is simply $1 - \exp[-\tau_S(r)]$.
We find that the Sobolev approximation applies for nearly all of the
models presented in this paper, and therefore provides a convenient
approach to estimating the optical depth.

@@

Using the methodology described in $\S$~\ref{sec:method}, we
propagated photons from the source and through the outflow to an
`observer' at $r \gg r_{\rm outer}$ who views the entire wind+source
complex.  Figure~\ref{fig:fiducial_1d} presents the 1D spectrum
that this observer would record, with the unattenuated flux
normalized to unit value.   The \ion{Mg}{2} doublet
shows the canonical `P-Cygni profile' that characterizes a continuum
source embedded within an outflow.  Strong absorption is evident at
$\delta v  \equiv (\lambda/\lambda_0 - 1) < -50 \mkms$ in both
transitions (with equivalent widths $W^{2796}_a =
3.0$\AA\ and $W^{2803}_a = 1.3$\AA) and each shows emission at
positive velocities.  For an isotropic and dust-free model, the
total equivalent width of the doublet must be zero,
i.e.\ every photon
absorbed eventually escapes the system, typically at lower
energy.  The wind simply shuffles the photons in frequency space.
A simple summation of the absorption and emission equivalent widths
(Table~\ref{tab:fiducial_EW}) confirms this expectation.

Focusing further on the \ion{Mg}{2} absorption, one notes that the profiles lie
well above zero intensity and have similar depth even though their $f\lambda$
values differ by a factor of two.  In standard absorption-line
analysis, this is 
the tell-tale signature of a `cloud' that has a high optical depth (i.e.\
is saturated) which only partially covers the emitting source
\citep[e.g.][]{rvs05a,hkp+10}.  Our fiducial wind model, however, 
{\it entirely covers the source}; the apparent partial covering must
be related to a different effect.
Figure~\ref{fig:noemiss} further emphasizes this point by comparing the 
absorption profiles from Figure~\ref{fig:fiducial_1d} against an
artificial model where no absorbed photons are 
re-emitted.   As expected from the
$\tau_{2796}$ profile (Figure~\ref{fig:fiducial_nvt}), this
`intrinsic' model
produces a strong \mgiid\ doublet that absorbs all photons at
$\delta v \approx -100 \mkms$, i.e.\ $I_{2796}^{\rm min} = \exp(-\tau^{\rm
  max}_{2796}) \approx 0$.
The true model, in contrast, has been `filled in' at $\delta v \approx -100
\mkms$ by photons scattered in the wind.  An
absorption-line analysis that ignores these effects
would (i) systematically underestimate the true optical
depth and/or (ii) falsely conclude that the wind partially covers the
source.  We will find that these are
generic results, even for wind models that include
dust and are not fully isotropic.

@@

Now consider the \ion{Fe}{2} transitions:
the bottom left panel of Figure~\ref{fig:fiducial_1d} covers the
majority of the \ion{Fe}{2} UV1 transitions and several are
shown in the velocity plot.  The line
profile for \feiib\ is very similar to the \mgiid\ doublet;
one observes strong absorption to negative velocities and strong
emission at $\mdv > 0 \mkms$ producing a characteristic P-Cygni profile. 
Splitting
the profile at $\mdv = -50 \mkms$, we measure an equivalent width
$W^{2600}_{\rm a} = 1.16$\AA\ in absorption and $W^{2600}_{\rm e} =
-0.94$\AA\ in emission (Table~\ref{tab:fiducial_EW}) for a total
equivalent width of $W^{2600}_{\rm TOT} \approx 0.22$\AA.  
In contrast, the \feiia\ resonance line shows much weaker emission and
a much higher total equivalent width ($W^{2586}_{\rm TOT} = 0.49$\AA),
even though the line has a $2 \times$ lower $f\lambda$ value.
These differences between the \ion{Fe}{2} resonance lines (and between
\ion{Fe}{2} and \ion{Mg}{2}) occur because of the complex of non-resonant
\feiis\ transitions that are coupled to the resonance lines
(Figure~\ref{fig:energy}).  Specifically, 
a resonance photon absorbed at \feiid\ has a finite probability of
being re-emitted as a non-resonant photon which then escapes the system
without further interaction.  The principal effects are to reduce the
line emission of \feiid\ and to produce non-resonant line-emission (e.g.\
\feiic).

@@

The preceding discussion emphasizes the filling-in of resonance
absorption at $\mdv
\lesssim -50 \mkms$ and the generation of emission lines at $\mdv \approx
0 \mkms$ by photons scattered in the wind.  We also
mapped the emission of the fiducial
model to sudy its spatial extent (see $\S$~\ref{sec:monte} for a description of the algorithm).
The output is a set of surface-brightness maps in a series of 
frequency channels yielding a
dataset analogous to integral-field-unit (IFU) observations.  In
Figure~\ref{fig:fiducial_ifu_mgii}, we present the output 
at several velocities relative to the \mgiia\
transition. At $\mdv = -250 \mkms$, where the wind has an optical
depth $\tau_{2796} < 1$ (Figure~\ref{fig:fiducial_nvt}),
the source contributes roughly half of the observed flux.  
At $\mdv=-100 \mkms$, however, the
wind absorbs all photons from the source and the observed emission is
entirely from photons scattered by the wind.  This scattered emission
actually exceeds the source+wind emission at 
$\mdv = -250 \mkms$ such that the absorption profile is
negatively offset from the velocity where $\tau_{2796}$ is maximal
(Figure~\ref{fig:fiducial_nvt}).
The net result is
weaker \ion{Mg}{2} absorption that peaks blueward of the actual peak in the
optical depth profile.  
Table~\ref{tab:line_diag} reports several kinematic measurements of
the absorption and emission features.
Clearly, these effects complicate estimates for the
speed, covering fraction, and total column density of the wind.  At
$\mdv = 0 \mkms$, the wind and source have comparable total flux with the
latter dominating at higher velocities.

Similar results are observed for the \ion{Fe}{2} resonance
transitions (Figure~\ref{fig:fiducial_ifu_feii}).
For transitions to fine-structure levels of the \aconfig\
configuration, the source
is unattenuated but there is a significant contribution from photons
generated in the wind.
